Plug-and-play dual-tree algorithm runtime analysis

### 2014-11-12 bounds.tex
Remove negative vspace and remove NIPS formatting.
Some thoughts on tree imbalance.
Update; rewritten intro, imbalance, and bound.

\documentclass{article} % For LaTeX2e
\usepackage{nips14submit_e,times}
\usepackage{hyperref}
\hypersetup{colorlinks,
            urlcolor=blue,
            citecolor=green,
            linkcolor=red}
\usepackage{url}
%\documentstyle[nips13submit_09,times,art10]{article} % For LaTeX 2.09
\usepackage{mathrsfs}
\usepackage{amsthm}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{algorithmic}
\usepackage{algorithm}
\usepackage{color}

@@

\title{Plug-and-play dual-tree algorithm runtime analysis}

@@

%\nipsfinalcopy % Uncomment for camera-ready version

@@

\vspace*{-0.8em}
\begin{abstract}
%A large collection of machine learning algorithms contain subproblems that can
%be efficiently solved with trees.  In the situation where a dual-tree algorithm
%with cover trees is used to solve a problem, rigorous worst-case runtime bounds
%can often be shown.  A worst-case runtime bound is proven for all cover tree
%dual-tree algorithms; this bound can be employed to give useful runtime bounds
%for individual problems.  We show a new, tighter bound of $O(c^9 \nu N)$ for
%all-nearest-neighbor search and also provide linear runtime bounds for
%approximate kernel density estimation and exact range count/search under some
%mild assumptions.  These are the first linear runtime bounds we are aware of for
%range count/search.
Numerous machine learning algorithms contain pairwise statistical
problems in their core. Often, tree structures are used
to solve these problems efficiently. Dual-tree
algorithms using cover trees have
rigorous worst-case runtime guarantees for some of these problems. In this paper, we present a {\em
problem-independent} runtime guarantee for {\em any} dual-tree
algorithm using the cover tree, separating out the problem-dependent and
the problem-independent elements. This allows us to just plug in
bounds for the problem-dependent elements to get runtime guarantees
for dual-tree algorithms for any pairwise statistical problem. We
demonstrate this plug-and-play procedure for nearest-neighbor search
and approximate kernel density estimation to get improved runtime
guarantees.  Under some assumptions, we also present the first linear runtime
guarantee for dual-tree based range search.
\end{abstract}

\vspace*{-1em}

%\section{Dual-tree algorithms for pairwise statistical problems (Introduction)}
\section{Dual-tree algorithms}
\vspace*{-0.5em}

%\begin{itemize}
%\item What are pairwise statistical problems? Examples?
%
%A surprising number of machine learning algorithms have bottlenecks which, if
%implemented naively, depend on the calculation of potentially large numbers of
%distances.  A common example is nearest neighbor search: if search
%for the nearest neighbor of a single query point in a reference set of size $N$,
%simple linear scan costs $O(N)$ time.  For many decades, spatial data structures
%have been employed to accelerate this type of computation, beginning with the
%quad tree \cite{finkel1974quad}; building trees on the reference set often
%results in expected runtimes of $O(\log N)$ per query point.
%
A surprising number of machine learning algorithms have computational
bottlenecks that can be expressed as pairwise statistical problems.
All-nearest-neighbor search is one such problem, naively requiring $O(N)$ time
to answer in single query in a reference set of size $N$; answering $O(N)$
queries subsequently requiring the prohibitive $O(N^2)$ time. Kernel density
estimation is also a pairwise statistical problem, naively requiring $O(N^2)$
time to answer $O(N)$ queries. The reference set is typically indexed with
spatial data structures to accelerate this type of computation
\cite{finkel1974quad, langford2006}; these result in $O(\log N)$ runtime per
query under favorable conditions.

@@

%\item General tree/problem independent dual-tree algorithms (maybe)
%\item What is lacking -- general dual-tree analysis (since the algorithms are so similar in structure)
%A recent framework for understanding and unifying dual-tree algorithms
%observes that all dual-tree algorithms are particularly similar, and generalizes
%them \cite{curtin2013tree}.  A
%nice benefit of this viewpoint is that it allows quick and easy development of
%new dual-tree algorithms.  In this paper, we make the same observation about
%runtime proofs for cover tree dual-tree algorithms, and show that
%runtime analysis of generalized cover tree dual-tree algorithms is
%possible.  Thus, the contributions of this paper are:
%
%\vspace*{-0.5em}
%\begin{itemize}
%  \item A general runtime bound for dual-tree algorithms that use the cover
%tree.
%
%  \item \mbox{A nearest neighbor search runtime bound,
%slightly improving on existing linear-time results.}
%
%  \item An approximate kernel density estimation runtime bound,
%improving on existing linear-time results.
%
%  \item Novel runtime analysis of dual-tree range count and range
%search.
%\end{itemize}
Curtin et.~al. \cite{curtin2013tree} observe that all dual-tree
algorithms are similar and propose a unified framework which allows the
development of dual-tree algorithms in a problem independent (and tree
independent) manner. In this paper, we make the same observation about runtime
proofs for cover tree dual-tree algorithms, and show that runtime analysis
of a generalized dual-tree algorithm is possible. The specific contributions of
this paper are:
\vspace*{-0.3em}
\begin{itemize}
  \item A general runtime bound for dual-tree algorithms that use the cover
tree.
  \item An improved linear runtime guarantee for all nearest-neighbor search.
  \item An improved linear runtime guarantee for approximate kernel density
estimation.
  \item The first linear runtime guarantee for dual-tree range count and range
search.
\end{itemize}

@@

\vspace*{-1em}
\section{Preliminaries}
\vspace*{-0.6em}

@@

\vspace*{-0.6em}
\section{General runtime bound}
\vspace*{-0.4em}

@@

\vspace*{-0.8em}
\section{Nearest neighbor search}
\vspace*{-0.4em}

@@

\vspace*{-0.6em}
\section{Approximate kernel density estimation}
\vspace*{-0.6em}

@@

\vspace*{-0.8em}
\section{Range Search and Range Count}
\vspace*{-0.4em}

@@

% Appendix doesn't have anything useful or non-straightforward.
%\appendix
%\input{appendix-ec}

>>>>

\documentclass{article} % For LaTeX2e
\usepackage{hyperref}
\hypersetup{colorlinks,
            urlcolor=blue,
            citecolor=green,
            linkcolor=red}
\usepackage{url}
%\documentstyle[nips13submit_09,times,art10]{article} % For LaTeX 2.09
\usepackage{mathrsfs}
\usepackage{amsthm}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{algorithmic}
\usepackage{algorithm}
\usepackage{color}

@@

\title{Plug-and-play dual-tree algorithm runtime analysis}
\author{Ryan R. Curtin, Dongryeol Lee, William B. March, Parikshit Ram}
\date{\today}

@@

@@

\begin{abstract}
%A large collection of machine learning algorithms contain subproblems that can
%be efficiently solved with trees.  In the situation where a dual-tree algorithm
%with cover trees is used to solve a problem, rigorous worst-case runtime bounds
%can often be shown.  A worst-case runtime bound is proven for all cover tree
%dual-tree algorithms; this bound can be employed to give useful runtime bounds
%for individual problems.  We show a new, tighter bound of $O(c^9 \nu N)$ for
%all-nearest-neighbor search and also provide linear runtime bounds for
%approximate kernel density estimation and exact range count/search under some
%mild assumptions.  These are the first linear runtime bounds we are aware of for
%range count/search.
Numerous machine learning algorithms contain pairwise statistical
problems in their core. Often, tree structures are used
to solve these problems efficiently. Dual-tree
algorithms using cover trees have
rigorous worst-case runtime guarantees for some of these problems. In this paper, we present a {\em
problem-independent} runtime guarantee for {\em any} dual-tree
algorithm using the cover tree, separating out the problem-dependent and
the problem-independent elements. This allows us to just plug in
bounds for the problem-dependent elements to get runtime guarantees
for dual-tree algorithms for any pairwise statistical problem. We
demonstrate this plug-and-play procedure for nearest-neighbor search
and approximate kernel density estimation to get improved runtime
guarantees.  Under some assumptions, we also present the first linear runtime
guarantee for dual-tree based range search.
\end{abstract}

%\section{Dual-tree algorithms for pairwise statistical problems (Introduction)}
\section{Dual-tree algorithms}

%\begin{itemize}
%\item What are pairwise statistical problems? Examples?
%
%A surprising number of machine learning algorithms have bottlenecks which, if
%implemented naively, depend on the calculation of potentially large numbers of
%distances.  A common example is nearest neighbor search: if search
%for the nearest neighbor of a single query point in a reference set of size $N$,
%simple linear scan costs $O(N)$ time.  For many decades, spatial data structures
%have been employed to accelerate this type of computation, beginning with the
%quad tree \cite{finkel1974quad}; building trees on the reference set often
%results in expected runtimes of $O(\log N)$ per query point.
%
A surprising number of machine learning algorithms have computational
bottlenecks that can be expressed as pairwise statistical problems.
Nearest neighbor search is one such problem, naively requiring $O(N)$ time
to answer in single query in a reference set of size $N$; answering $O(N)$
queries subsequently requires prohibitive $O(N^2)$ time. Kernel density
estimation is also a pairwise statistical problem, naively requiring $O(N^2)$
time to answer $O(N)$ queries. The reference set is typically indexed with
spatial data structures to accelerate this type of computation
\cite{finkel1974quad, langford2006}; these result in $O(\log N)$ runtime per
query under favorable conditions.

@@

Gray and Moore \cite{nbody} generalized the fast multipole method from
computational physics to obtain dual-tree algorithms.  These are extremely
useful when there are large query sets, not just a few query points.  Dual-tree
algorithms can be easily understood through the recent framework of Curtin
et~al. \cite{curtin2013tree}: two trees (a query tree and a reference tree) are
traversed by a {\it pruning dual-tree traversal}.  This traversal visits
combinations of nodes from the trees in some sequence (each combination consisting of a query node and a
reference node), calling a problem-specific \texttt{Score()} function to
determine if the node combination can be pruned.  If not, then a
problem-specific \texttt{BaseCase()} function is called for each combination of
points held in the query node and reference node.  This has significant
similarity to the more common single-tree branch-and-bound algorithms, except
that the algroithm must recurse into child nodes of {\it both} the query tree
and reference tree.

There exist numerous dual-tree algorithms for problems as diverse as kernel
density estimation \cite{gray2003nonparametric}, mean shift \cite{wang2007fast},
minimum spanning tree calcul

%\item General tree/problem independent dual-tree algorithms (maybe)
%\item What is lacking -- general dual-tree analysis (since the algorithms are so similar in structure)
%A recent framework for understanding and unifying dual-tree algorithms
%observes that all dual-tree algorithms are particularly similar, and generalizes
%them \cite{curtin2013tree}.  A
%nice benefit of this viewpoint is that it allows quick and easy development of
%new dual-tree algorithms.  In this paper, we make the same observation about
%runtime proofs for cover tree dual-tree algorithms, and show that
%runtime analysis of generalized cover tree dual-tree algorithms is
%possible.  Thus, the contributions of this paper are:
%
%\begin{itemize}
%  \item A general runtime bound for dual-tree algorithms that use the cover
%tree.
%
%  \item \mbox{A nearest neighbor search runtime bound,
%slightly improving on existing linear-time results.}
%
%  \item An approximate kernel density estimation runtime bound,
%improving on existing linear-time results.
%
%  \item Novel runtime analysis of dual-tree range count and range
%search.
%\end{itemize}
Curtin et.~al. \cite{curtin2013tree} observe that all dual-tree
algorithms are similar and propose a unified framework which allows the
development of dual-tree algorithms in a problem independent (and tree
independent) manner. In this paper, we make the same observation about runtime
proofs for cover tree dual-tree algorithms, and show that runtime analysis
of a generalized dual-tree algorithm is possible. The specific contributions of
this paper are:
\begin{itemize}
  \item A general runtime bound for dual-tree algorithms that use the cover
tree.
  \item An improved linear runtime guarantee for all nearest-neighbor search.
  \item An improved linear runtime guarantee for approximate kernel density
estimation.
  \item The first linear runtime guarantee for dual-tree range count and range
search.
\end{itemize}

@@

\section{Preliminaries}

@@

\section{Tree Imbalance}

\input{bounds-files/imbalance.tex}

\section{General runtime bound}

@@

\section{Nearest neighbor search}

@@

\section{Approximate kernel density estimation}

@@

\section{Range Search and Range Count}

@@

% Appendix doesn't have anything useful or non-straightforward.
\appendix
\input{appendix-ec}
